---
title: "Statlaer_comp_ex_3_task_3"
author: "Julie"
date: "30 april 2018"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results="hold",message = FALSE,warning=FALSE)
```



# Problem 3 - Unsupervised methods

## a) Principal component analysis 

* Q16. Explain what you see in the `biplot` in relation to the loadings for the first two principal components.

The loadings for the principal components are the  weights in front of the variables which gives the principal components:

$$
i^{th} \text{principal component}=Z_i=\sum_{p=1}^n\phi_{pi}X_p
$$
For the first principal component, wine and tea are the most significant factors with loadings of respectively -0.5 and 0.66. They point in each direction, which means they are negatively correlated. The other beverages are located closer to the center. As for liquer, it is almost not explained by PC1 at all, with a loading of 0.021.
With loadings of 0.67 and 0.58, coffee and cocoa are the most significant factors for the second principal component. For wine and tea, almost no information is explained by PC2. 
Loading vectors pointing in the same direction, provides information about the drinking habits. For example, a low or high consume of cocoa, tells us that there is a low or high consume of beer as well. 


* Q17. Does this analysis give you any insight into the consumption of beverages and similarities between countries?

From the correlation plot, we see what beverages are correlated. First of all, the most noticeable effect is the negative correlation between consumption of coffee and tea. There is also a negative correlation between consumption of tea and wine as well. In general, the correlation factors are quite low, with a highest factor in absolute value of correlation of 0.4. 

Between countries, we see that Great Britain and Ireland, which are closely located in the plot, but also geographically, have the same drinking habits and opinions about tea. Countries like Italy, France and Portugal have negative scores on the first component, with opposite meaning of drinking tea. Also, the loading vector for tea is longer than the one for liquer, which means the opinions about tea is stronger for countries as Great Britain and Ireland, than for Hungary and the Soviet Union. 

Since the Netherlands has a large positive score on the second component, we can assume that they have very strong opinions about cocoa.

At last, we can not by looking directly at the biplot say if countries like or dislike a given beverage, since they only are unique up to a sign flip. But using the fact that we know that Italy, Spain and Portugal produce a lot of wine, we can assume that they consume a lot, and that Great Britain and Ireland consume a lot of tea.
```{r,echo=FALSE,eval=FALSE}
# reading data on consumption of different beverages for countries
drink <- read.csv("https://www.math.ntnu.no/emner/TMA4267/2017v/drikke.TXT",sep=",",header=TRUE)
drink <- na.omit(drink)
# looking at correlation between consumptions
drinkcorr=cor(drink)
library(corrplot)
corrplot(drinkcorr,method="circle")
# now for PCA
pcaS <- prcomp(drink,scale=TRUE) # scale: variables are scaled 
pcaS$rotation
summary(pcaS)
biplot(pcaS,scale=0,cex=0.6) # scale=0: arrows scaled to represent the loadings
```

## b) Hierarchical clustering 

```{r,echo=FALSE,eval=FALSE}
library(knitr)
species=c("Human","Chimpanzee","Gorilla","Orangutan","Gibbon")
distJC <- matrix(c(0,1,3,9,12,
                   1,0,2,8,11,
                   3,2,0,6,11,
                   9,8,6,0,11,
                   12,11,11,11,0),5,5)
dimnames(distJC) <- list(species,species)
kable(distJC)
d=as.dist(distJC)
h_a=hclust(d,method="average")
h_c=hclust(d,method="complete")
h_s=hclust(d,method="single")
plot(h_c)
plot(h_a)
plot(h_s)

h_c$height
h_a$height
h_s$height

```


* Q18. Describe how the distance between _clusters_ are defined for single, complete and average linkage.

Linkage is defined as the dissimilarity between groups of observations. 
Single: minimal intercluster dissimilarity. The smallest dissimilarity between cluster A and B is  computed. 
Complete: Maximal intercluster dissimilarity. The largest dissimilarity between A and B  is recorded.
Average: The average intercluser dissimilarity. The average dissimilarity between A and B is recorded.

* Q19. Identify which of the three dendrograms (A, B, C) correspond to the three methods single, complete and average linkage. Justify your solution.

All the cluster dendograms has height 1 between human and chimpanze.
Dendrogram A has heights 1, 3, 9 and 12. These are the largest dissimilarities between human and the different animals, as in column 1 in the distance matrix, which means that A corresponds to complete linkage.
Dendrogram B has heights 1, 2.50, 7.67 and 11.25. Between Gorilla and the cluster (Human/Chimpanzee), the distance is 2.5, which corresponds to $\frac{1}{2}(G-H+G-C)=\frac{5}{2}=2.5$. The same reasoning can be used for the other heights: 
$\frac{1}{3}(O-H+O-C+O-G)=\frac{23}{3}=7.67$ for two clusters and  $\frac{1}{4}(Gibb-H+Gibb-C+Gibb-G+Gibb-O)=\frac{45}{4}=11.25$. This way of averaging is called weighted pair group method average, and differs from unweighted pair group method average, which only looks at the average between the clusters.
We can therefore conclude that dendrogram B corresponds to average linkage. 
The last dendrogram, C, has heights 1,2,6 and 11, which corresponds to single linkage because the shortest distance between clusters is computed. 
