---
title: "Statlaer_comp_ex_3_task_3"
author: "Julie"
date: "30 april 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results="hold",message = FALSE,warning=FALSE)
```




# Problem 3 - Unsupervised methods

## a) Principal component analysis 

* Q16. Explain what you see in the `biplot` in relation to the loadings for the first two principal components. 
* Q17. Does this analysis give you any insight into the consumption of beverages and similarities between countries? 

```{r,echo=FALSE,eval=FALSE}
# reading data on consumption of different beverages for countries
drink <- read.csv("https://www.math.ntnu.no/emner/TMA4267/2017v/drikke.TXT",sep=",",header=TRUE)
drink <- na.omit(drink)
# looking at correlation between consumptions
drinkcorr=cor(drink)
library(corrplot)
corrplot(drinkcorr,method="circle")
# now for PCA
pcaS <- prcomp(drink,scale=TRUE) # scale: variables are scaled 
pcaS$rotation
summary(pcaS)
biplot(pcaS,scale=0,cex=0.6) # scale=0: arrows scaled to represent the loadings
```

## b) Hierarchical clustering 

* Q18. Describe how the distance between _clusters_ are defined for single, complete and average linkage. 
* Q19. Identify which of the three dendrograms (A, B, C) correspond to the three methods single, complete and average linkage. Justify your solution.