---
title: "comp_ex_3_task4.Rmd"
author: "Julie"
date: "May 2, 2018"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 4 - Neural networks

```{r}
library(keras)
imdb <- dataset_imdb(num_words = 10000)

train_data <- imdb$train$x
train_labels <- imdb$train$y
test_data <- imdb$test$x
test_labels <- imdb$test$y

vectorize_sequences <- function(sequences, dimension = 10000) {
  results <- matrix(0, nrow = length(sequences), ncol = dimension)
  for (i in 1:length(sequences))
    results[i, sequences[[i]]] <- 1                                     
  results
}

x_train <- vectorize_sequences(train_data)
x_test <- vectorize_sequences(test_data)
y_train <- as.numeric(train_labels)
y_test <- as.numeric(test_labels)


```


Use 4 units instead of 16: 
```{r}
model <- keras_model_sequential() %>%
  layer_dense(units = 32, activation = "relu", input_shape = c(10000)) %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")

model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)
```

```{r}
val_indices <- 1:10000

x_val <- x_train[val_indices,]
partial_x_train <- x_train[-val_indices,]
y_val <- y_train[val_indices]
partial_y_train <- y_train[-val_indices]

```

```{r}
history <- model %>% fit(
  partial_x_train,
  partial_y_train,
  epochs = 5,
  batch_size = 512,
  validation_data = list(x_val, y_val)
)
```


```{r}
plot(history)
history_df <- as.data.frame(history)
min(history_df$value)
```


* Q20. What is the advantage of using a non-linear activation function such as `relu`?

The advantage is that we get a richer hypothesis space, because if not, the network would only learn linear transformations of the input data, and not take advantage of the possibility of having many layers.

* Q21. Why do we need to use a different activation function (`sigmoid`) in the output layer instead of using `relu` again? 

We want the output to be represented as a probability distribution, and then all nodes in the output layer have to take a value between 0 and 1. `relu`can be greater than 1, while `sigmoid` only returns values between 0 and 1. 

* Q22. Plot the training and validation loss and accuracy for the simpler and more complex model mentioned above. How do they compare with the model with 16 hidden units?




* Q23. Besides reducing the network's size, what other methods can be used to avoid overfitting with neural network models? Briefly describe the intuition behind each one.

- Get more data: It is important to have more data than the number of effective parameters. Therefor, getting more data will help reducing overfitting.
- Add weight reqularization: Since we optimize our loss function, we want to penalize high weight functions. WHY?
- Add dropout: Adding dropout means adding noise to the model, so the network not will memorize patterns that are not significant. 
